---
title: "NGES Consulting Engagement - Day 2: Predictive Modeling, Infrastructure Risk, Demand Response & ROI"
author: "Ujwal Dasari"
date: "`r format(Sys.time(), '%B %d, %Y  %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
    number_sections: true
  word_document:
    toc: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5)
c_blue   <- "#2a78d6"; c_orange <- "#eb6834"; c_aqua <- "#1baf7a"
c_red    <- "#d03b3b"; c_muted  <- "#898781"; c_grid <- "#e1e0d9"
theme_nges <- function() {
  ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(panel.grid.minor = ggplot2::element_blank(),
                   panel.grid.major = ggplot2::element_line(color = c_grid),
                   plot.title = ggplot2::element_text(face = "bold"),
                   axis.text = ggplot2::element_text(color = c_muted))
}
```

# Introduction

Day 1 told us where the signal is: demand tracks customer segment loads almost
perfectly and ignores the weather, infrastructure risk is concentrated in a
31-substation critical tier, and demand response results swing wildly per dollar
spent. Day 2 turns those observations into the four analytical deliverables the
client asked for: a forecasting model comparison, an infrastructure investment
priority list, a demand response optimization case, and a financial impact and
ROI assessment. Everything here feeds Sunday's Board presentation and the
Power BI dashboard.

```{r load}
library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
# ranger is a fast random forest implementation; install once if needed:
# install.packages("ranger")
library(ranger)

data_dir <- if (file.exists("NGES Energy Demand Forecasting Dataset.csv")) "." else "../data"
demand <- read_csv(file.path(data_dir, "NGES Energy Demand Forecasting Dataset.csv"))
infra  <- read_csv(file.path(data_dir, "NGES Infrastructure Risk & Planning Dataset.csv"))
dr     <- read_csv(file.path(data_dir, "NGES Demand Response Effectiveness Dataset.csv"))

demand <- demand %>%
  mutate(Timestamp = mdy_hm(Timestamp),
         Hour  = hour(Timestamp),
         Month = month(Timestamp),
         Dow   = wday(Timestamp)) %>%
  arrange(Timestamp)
```

# Feature Engineering & Evaluation Design

```{r features}
# Median-impute the ~2% missing weather readings documented on Day 1
demand <- demand %>%
  mutate(Temperature = replace_na(Temperature, median(Temperature, na.rm = TRUE)),
         Humidity    = replace_na(Humidity,    median(Humidity,    na.rm = TRUE)))

# Time-based split: first 80% of hours train the models, the most recent 20%
# (roughly the 2025 portion) tests them. A random split would leak the future
# into training and flatter every model, so we avoid it.
cut_row <- floor(nrow(demand) * 0.8)
train <- demand[1:cut_row, ]
test  <- demand[(cut_row + 1):nrow(demand), ]
range(train$Timestamp); range(test$Timestamp)

# Helper to score any set of predictions on the test window
score <- function(actual, pred) {
  tibble(MAE  = round(mean(abs(actual - pred)), 1),
         RMSE = round(sqrt(mean((actual - pred)^2)), 1),
         R2   = round(1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2), 3))
}
```

We evaluate every model on the same held-out window and against the same
yardstick: the client's own forecast, which scores an MAE of about 24.0 MW on
that window. A model only matters if it beats that number.

# Baseline Model: Weather and Calendar Only

```{r baseline}
# Linear regression using only the variables Theory 1 says should matter:
# weather, time of day, month, day of week, region and customer count.
lm_weather <- lm(ActualDemandMW ~ Temperature + Humidity + WindSpeed +
                   factor(Hour) + factor(Month) + factor(Dow) +
                   Region + CustomerCount, data = train)
score(test$ActualDemandMW, predict(lm_weather, test)) %>%
  kable(caption = "Baseline: weather + calendar linear model")
```

The baseline fails completely, and that failure is informative. An MAE around
113 MW and an R-squared of essentially zero means weather and calendar carry no
usable forecasting signal in this data. Any investment aimed at "better weather
forecasting" would not improve demand forecasts at NGES.

# Segment-Load Models

```{r load-models}
# Same linear model, now adding the three customer segment loads
lm_loads <- lm(ActualDemandMW ~ ResidentialLoadMW + CommercialLoadMW +
                 IndustrialLoadMW + Temperature + Humidity + WindSpeed +
                 factor(Hour) + factor(Month) + Region, data = train)
res_lm <- score(test$ActualDemandMW, predict(lm_loads, test))

# Advanced model: random forest on the full feature set
rf <- ranger(ActualDemandMW ~ ResidentialLoadMW + CommercialLoadMW +
               IndustrialLoadMW + Temperature + Humidity + WindSpeed +
               Hour + Month + Dow + CustomerCount + Region,
             data = train, num.trees = 300, importance = "impurity", seed = 42)
res_rf <- score(test$ActualDemandMW, predict(rf, test)$predictions)

# The client's own forecast on the identical test window, as the benchmark
bench <- test %>% filter(!is.na(ForecastedDemandMW))
res_nges <- score(bench$ActualDemandMW, bench$ForecastedDemandMW)

bind_rows("Current NGES forecast"      = res_nges,
          "Linear model + segment loads" = res_lm,
          "Random forest (all features)" = res_rf,
          .id = "Model") %>%
  kable(caption = "Model comparison on the 2025 hold-out window")
```

The segment-load linear model beats the incumbent forecast clearly, landing
near an MAE of 20 MW (R-squared about 0.97) against the current forecast's
24 MW, roughly a 17% error reduction. The random forest, despite its extra
machinery, only matches the incumbent at about 24 MW; exact values shift
slightly with the random seed. It is worth pausing on the fact that the simple,
fully explainable linear model wins: demand here is close to the sum of its
segment loads plus irreducible noise of about 25 MW, so extra model complexity
buys nothing and here even costs a little. For a regulated utility that must
defend its planning methods to a public utility commission, an interpretable
model that outperforms is the best possible outcome.

```{r importance}
# Which variables does the random forest actually lean on?
enframe(ranger::importance(rf), name = "Feature", value = "Importance") %>%
  mutate(Share = round(Importance / sum(Importance) * 100, 1)) %>%
  arrange(desc(Share)) %>%
  dplyr::select(Feature, Share) %>%
  head(8) %>%
  ggplot(aes(reorder(Feature, Share), Share)) +
  geom_col(fill = c_blue, width = 0.6) +
  geom_text(aes(label = paste0(Share, "%")), hjust = -0.15, size = 3.5) +
  coord_flip() +
  labs(title = "Random forest feature importance",
       x = NULL, y = "Share of importance (%)") +
  theme_nges()
```

Residential load alone carries roughly half the model's importance, commercial
about a quarter and industrial about a fifth, with every weather and calendar
variable under 1%. The practical translation: NGES's forecasting future is
segment-level load forecasting from AMI meter data, refreshed near real time,
not weather modeling.

## Predicting Peak Events

```{r peaks}
# Day 1 showed peak events are exactly the top decile of demand, so an accurate
# demand forecast doubles as a peak-event early-warning system. We flag any hour
# whose predicted demand crosses the historical peak threshold.
thresh <- min(train$ActualDemandMW[train$PeakDemandEvent == 1])
pred_peak <- as.integer(predict(lm_loads, test) >= thresh)
actual_peak <- test$PeakDemandEvent
tibble(Precision = round(sum(pred_peak & actual_peak) / sum(pred_peak), 3),
       Recall    = round(sum(pred_peak & actual_peak) / sum(actual_peak), 3)) %>%
  kable(caption = "Peak-event detection from the demand forecast")
```

Flagging hours whose forecast crosses the peak threshold catches the large
majority of true peak events with high precision. Hours near the boundary are
the misses, which is exactly where a plus-or-minus 20 MW error band bites. In
operations this becomes a simple traffic-light rule: forecast above the
threshold, or within one error band of it, triggers demand response
pre-positioning.

# Infrastructure Risk Analysis & Capital Priorities

```{r infra-tiers}
# Tier the 500 substations by composite risk score. We use explicit >= boundaries
# (rather than cut) so a score of exactly 70 lands in the critical tier.
infra <- infra %>%
  mutate(RiskTier = factor(case_when(RiskScore >= 70 ~ "Critical",
                                     RiskScore >= 55 ~ "Elevated",
                                     RiskScore >= 40 ~ "Moderate",
                                     TRUE            ~ "Low"),
                           levels = c("Low", "Moderate", "Elevated", "Critical")))
infra %>%
  count(RiskTier, name = "Substations") %>%
  mutate(Share_Pct = round(Substations / sum(Substations) * 100, 1)) %>%
  kable(caption = "Substations by risk tier")

# Regional distribution of the critical tier
infra %>%
  filter(RiskTier == "Critical") %>%
  count(Region, name = "CriticalSubstations") %>%
  arrange(desc(CriticalSubstations)) %>%
  kable(caption = "Critical-tier substations by region")
```

Thirty-one substations (6.2%) sit in the critical tier, concentrated in the
North (10) and West (8). Another 35 assets systemwide are running past 100% of
rated capacity, meaning they depend on favorable conditions every single day.

```{r priority-list}
# Investment priority: rank critical-tier assets by a blend of risk score and
# projected load growth, since growth converts today's risk into tomorrow's outage.
priority <- infra %>%
  filter(RiskTier == "Critical") %>%
  mutate(PriorityScore = round(0.7 * RiskScore + 3 * ProjectedLoadGrowthPct, 1)) %>%
  arrange(desc(PriorityScore)) %>%
  dplyr::select(SubstationID, Region, InfrastructureAgeYears, CurrentUtilizationPct,
         OutagesLast5Years, ProjectedLoadGrowthPct, RiskScore, PriorityScore)
head(priority, 15) %>% kable(caption = "Top 15 capital investment priorities")

# Visual: utilization vs risk with the critical tier highlighted
ggplot(infra, aes(CurrentUtilizationPct, RiskScore)) +
  geom_point(aes(color = RiskTier == "Critical"), size = 1.6, alpha = 0.6) +
  scale_color_manual(values = c(`FALSE` = c_blue, `TRUE` = c_red),
                     labels = c("Other substations", "Critical tier"), name = NULL) +
  geom_vline(xintercept = 100, linetype = "dashed", color = c_muted) +
  labs(title = "Utilization vs risk score, critical tier in red",
       x = "Current utilization (%)", y = "Composite risk score") +
  theme_nges()
```

The recommendation that falls out of this table is phased and affordable: put
the 31 critical-tier assets on a 3-5 year upgrade schedule ordered by the
priority score, starting with substations like 1445 (West, 48 years old, 92.8%
utilized) and 1402/1434 (North, both past 92% with double-digit projected
growth). The roughly 130 elevated-tier assets get condition monitoring and
demand response coverage instead of immediate capital, which is what funds the
deferral value in the ROI section. One caution for the Board: substation 1103
tops the raw risk list mostly on outage history (8 in five years) despite modest
utilization, a reminder the composite score blends several failure modes and
engineering review should confirm each project before commitment.

# Demand Response Optimization

```{r dr-quartiles}
dr <- dr %>% mutate(CostPerMW = ProgramCost / DemandReductionMW)

# Split the 8,000 events into cost-effectiveness quartiles
dr <- dr %>% mutate(Quartile = ntile(CostPerMW, 4))
dr %>%
  group_by(Quartile) %>%
  summarise(Events        = n(),
            AvgCostPerMW  = round(mean(CostPerMW), 0),
            TotalCost_M   = round(sum(ProgramCost) / 1e6, 1),
            TotalMW       = round(sum(DemandReductionMW), 0),
            .groups = "drop") %>%
  kable(caption = "DR events by cost-effectiveness quartile")
```

The quartile table is the whole demand response story in four rows. The best
quartile of events removes load at a little over $100 per MW; the worst quartile
averages well over $1,100 per MW, eleven times more, and absorbed $141.3 million
of the program's $420.2 million total spend while delivering only about 10% of
the megawatts. Because Day 1 showed customer types perform identically, this is
an execution problem (event timing, sizing and incentive design), not a
recruitment problem.

```{r dr-savings}
# Savings case: re-run the worst quartile's MW at the program's own median cost
med_cpm <- median(dr$CostPerMW)
worst   <- dr %>% filter(Quartile == 4)
savings <- sum(worst$ProgramCost) - sum(worst$DemandReductionMW) * med_cpm
tibble(MedianCostPerMW   = round(med_cpm, 0),
       WorstQuartileCost_M = round(sum(worst$ProgramCost) / 1e6, 1),
       SameMWatMedian_M    = round(sum(worst$DemandReductionMW) * med_cpm / 1e6, 1),
       Savings_M           = round(savings / 1e6, 1)) %>%
  kable(caption = "Savings if the worst quartile performed at the median")
```

Delivering the worst quartile's megawatts at the program's own median
cost-effectiveness ($330/MW, a standard the program already meets half the
time) would have cost $42.8 million instead of $141.3 million, freeing $98.5
million over the period with zero megawatts of peak reduction given up. Even a
narrower fix limited to the 1,075 events that cost more than $1,000/MW recovers
about $62.6 million. Notably, these figures bracket the $95-120 million exposure
Executive Leadership quoted in the engagement letter: most of that money is
recoverable from inside the DR budget alone.

# Financial Impact & ROI

```{r roi}
# The dataset spans roughly 2.83 years, so period savings are annualized by /2.83
roi <- tribble(
  ~Initiative,                      ~Conservative, ~Expected, ~Optimistic,
  "DR optimization ($M/yr)",         22.1,          34.8,      40.0,
  "Risk-based capital targeting ($M/yr)", 6.0,      10.0,      15.0,
  "Forecast upgrade ($M/yr)",        0.5,           1.0,       2.0)
roi %>%
  bind_rows(summarise(roi, Initiative = "Total annual value",
                      across(where(is.numeric), sum))) %>%
  kable(caption = "Estimated annual value by scenario")
```

The assumptions behind each line, stated plainly so the Board can challenge them:

- **DR optimization.** The $98.5M period saving annualizes to $34.8M (expected).
  Conservative takes only the over-$1,000/MW fix ($62.6M, or $22.1M/yr);
  optimistic adds modest procurement leverage on the remaining events.
- **Risk-based capital targeting.** Concentrating near-term capital on 31
  critical assets instead of a broad age-based program defers roughly $100-150M
  of lower-priority spend by 3-5 years; at a utility cost of capital near 7-8%,
  deferral plus avoided outage cost is worth an estimated $6-15M per year.
- **Forecast upgrade.** Cutting peak-hour forecast error about 17% trims
  emergency power purchases across roughly 880 peak hours per year; at a
  $100-200/MWh emergency premium this is worth $0.5-2M per year. It is the
  smallest line but also the cheapest to implement, and it is what makes the
  peak-warning system work.

Implementation costs are estimated at about $8M one-time (forecasting platform,
DR program redesign, analytics staffing) plus $5M per year to run. Against the
expected $45.8M of annual value, the portfolio pays back in well under a year
and returns roughly five to nine times its cost annually thereafter. Over a
three-year capital planning window the expected case creates about $137M of
value, which more than neutralizes the $95-120M exposure leadership identified.

# Limitations

Honesty about limits keeps these numbers defensible. The demand data contains
about 25 MW of hour-to-hour noise no model can remove, so forecast gains beyond
roughly 20 MW MAE should not be promised. Weather showed no effect in this
dataset; in most real service territories it does, so the finding should be
re-tested once NGES validates its weather feeds (the impossible humidity
readings suggest sensor problems). The demand file's substations do not join
cleanly to the infrastructure file, keeping the analysis at regional rather than
asset level. And the deferral and emergency-premium values rest on planning
assumptions the finance team should replace with NGES's actual cost of capital
and procurement records before regulatory filings.

# Conclusions

Three moves, in priority order: fix demand response execution (largest, fastest,
self-funding), target capital at the 31-substation critical tier (defers major
spend while cutting outage risk where it is concentrated), and upgrade
forecasting to a segment-load model (small direct savings, but it powers the
peak-event warning system and the dashboard). Together they turn a $95-120M
exposure into an expected $45.8M of annual value at a cost of about $5M per year
to sustain.
