---
title: "NGES Consulting Engagement - Day 1: Data Quality Review & Exploratory Data Analysis"
author: "Ujwal Dasari"
date: "`r format(Sys.time(), '%B %d, %Y  %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
    number_sections: true
  word_document:
    toc: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5)
# Chart palette used across the whole engagement so every figure looks consistent
c_blue   <- "#2a78d6"   # primary series color
c_orange <- "#eb6834"   # second series color
c_aqua   <- "#1baf7a"   # third series color
c_red    <- "#d03b3b"   # reserved for risk / problem highlights only
c_muted  <- "#898781"   # axis text and de-emphasized bars
c_grid   <- "#e1e0d9"   # hairline gridlines
theme_nges <- function() {
  ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(panel.grid.minor = ggplot2::element_blank(),
                   panel.grid.major = ggplot2::element_line(color = c_grid),
                   plot.title = ggplot2::element_text(face = "bold"),
                   axis.text = ggplot2::element_text(color = c_muted))
}
```

# Introduction

NorthGrid Energy Services (NGES) asked our consulting team to figure out why peak
demand events keep climbing (14 in 2022 up to 34 in 2025), which parts of the grid
carry the most risk, and whether their demand response spending is paying off.
Leadership put the financial exposure from demand volatility at $95-120 million.

Day 1 is about getting to know the data honestly before we model anything. We were
given three datasets: an hourly demand and forecasting file, a substation-level
infrastructure risk file, and a demand response event file. This report covers the
data quality checks and the exploratory analysis for all three, and closes with the
working hypotheses we will test with predictive models on Day 2.

```{r load-data}
# Load the tidyverse for data wrangling and plotting
library(tidyverse)
library(scales)
library(knitr)

# The three CSVs are expected either next to this .Rmd or two folders up
# (the Deliverables/Friday location inside the project folder).
data_dir <- if (file.exists("NGES Energy Demand Forecasting Dataset.csv")) "." else "../data"

# read_csv handles the byte-order mark at the start of these files automatically
demand <- read_csv(file.path(data_dir, "NGES Energy Demand Forecasting Dataset.csv"))
infra  <- read_csv(file.path(data_dir, "NGES Infrastructure Risk & Planning Dataset.csv"))
dr     <- read_csv(file.path(data_dir, "NGES Demand Response Effectiveness Dataset.csv"))

# Parse the timestamp (month/day/year hour:minute) and add calendar helpers
demand <- demand %>%
  mutate(Timestamp = mdy_hm(Timestamp),
         Date      = as_date(Timestamp),
         Hour      = hour(Timestamp),
         Month     = month(Timestamp),
         MonthLab  = floor_date(Timestamp, "month"),
         Weekend   = wday(Timestamp) %in% c(1, 7),
         Season    = case_when(Month %in% c(12, 1, 2) ~ "Winter",
                               Month %in% 3:5  ~ "Spring",
                               Month %in% 6:8  ~ "Summer",
                               TRUE            ~ "Fall"))
dim(demand); dim(infra); dim(dr)
```

The demand file holds 25,000 hourly records running from January 1, 2023 to
November 7, 2025 across five regions (North, South, East, West, Central) and 100
substation IDs. The infrastructure file covers 500 substations, and the demand
response file logs 8,000 program events.

# Data Quality Assessment

```{r quality}
# Count missing values per column in each dataset
colSums(is.na(demand)) %>% .[. > 0]
colSums(is.na(infra))  %>% sum()
colSums(is.na(dr))     %>% sum()

# Check for duplicate rows in all three files
c(demand = sum(duplicated(demand)),
  infra  = sum(duplicated(infra)),
  dr     = sum(duplicated(dr)))

# Sanity-check the physical ranges of the weather readings
summary(dplyr::select(demand, Temperature, Humidity, WindSpeed))
```

Three columns in the demand file have missing values: Temperature (500 rows),
Humidity (500 rows) and ForecastedDemandMW (500 rows), each about 2% of the file.
Only 6 rows are missing both a temperature and a forecast, so the gaps look
independent rather than caused by one broken feed. This matches the "missing
weather observations" warning in the engagement packet. The infrastructure and
demand response files are complete, and none of the three files contain duplicate
rows.

The weather columns do contain physically doubtful readings. Temperature runs as
high as 131.2 F and humidity ranges from -34.3 to 130.2, and negative humidity is
not possible. We treat these as sensor or data-entry errors. Because (as the next
sections show) weather turns out to have almost no relationship with demand in
this data, these outliers do not change any conclusion, but we document them and
will use median imputation for the missing weather values in the Day 2 models.

```{r error-check}
# Confirm the ForecastErrorMW column really equals Actual minus Forecast
demand %>%
  filter(!is.na(ForecastedDemandMW)) %>%
  summarise(max_inconsistency = max(abs(ForecastErrorMW -
                                    (ActualDemandMW - ForecastedDemandMW))))
```

The stored forecast error column reconciles exactly with actual minus forecast,
so the file is internally consistent.

# Demand Patterns Over Time

```{r monthly-trend, fig.cap="Average demand by month"}
# Average demand per calendar month across the whole period
monthly <- demand %>%
  group_by(MonthLab) %>%
  summarise(AvgDemand = mean(ActualDemandMW), .groups = "drop")

ggplot(monthly, aes(MonthLab, AvgDemand)) +
  geom_line(color = c_blue, linewidth = 1) +
  labs(title = "Average hourly demand by month, Jan 2023 - Nov 2025",
       x = NULL, y = "Average demand (MW)") +
  theme_nges()
```

The first surprise of the engagement shows up here. Monthly average demand is
almost flat, moving only between about 746 and 757 MW across nearly three years.
There is no summer air-conditioning hump and no winter heating spike. Whatever is
driving NGES's peak events, it is not the seasonal cycle we normally expect from
a utility.

```{r hourly-profile}
# Average demand by hour of day, and weekday versus weekend
demand %>%
  group_by(Hour) %>%
  summarise(AvgDemand = mean(ActualDemandMW), .groups = "drop") %>%
  ggplot(aes(Hour, AvgDemand)) +
  geom_line(color = c_blue, linewidth = 1) +
  labs(title = "Average demand by hour of day",
       x = "Hour", y = "Average demand (MW)") +
  theme_nges()

demand %>%
  group_by(Weekend) %>%
  summarise(AvgDemand_MW = round(mean(ActualDemandMW), 1), .groups = "drop") %>%
  kable(caption = "Weekday vs weekend demand")
```

The hourly profile is equally flat: every hour of the day averages within about
2 MW of every other hour, and weekends (749.9 MW) look the same as weekdays
(750.6 MW). Regional averages sit in a narrow band too, from 748.1 MW in the East
to 753.9 MW in the North. Calendar and geography explain very little here.

# Weather Sensitivity

```{r weather, fig.cap="Temperature vs demand"}
# Scatter a random sample of hours so the plot stays readable
set.seed(42)
demand %>%
  filter(!is.na(Temperature)) %>%
  slice_sample(n = 3000) %>%
  ggplot(aes(Temperature, ActualDemandMW)) +
  geom_point(color = c_aqua, alpha = 0.35, size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
  labs(title = "Temperature vs hourly demand (3,000 sampled hours)",
       x = "Temperature (F)", y = "Demand (MW)") +
  theme_nges()

# Correlation and peak-event rates by temperature band
cor(demand$Temperature, demand$ActualDemandMW, use = "complete.obs")

demand %>%
  filter(!is.na(Temperature)) %>%
  mutate(TempBand = cut(Temperature, c(-100, 32, 60, 80, 95, 200),
                        labels = c("Below 32F", "32-60F", "60-80F", "80-95F", "Above 95F"))) %>%
  group_by(TempBand) %>%
  summarise(Hours = n(),
            PeakEventRate_Pct = round(mean(PeakDemandEvent) * 100, 1),
            .groups = "drop") %>%
  kable(caption = "Peak-event rate by temperature band")
```

This is the most important early finding, because it contradicts the client's
Theory 1. The correlation between temperature and demand is 0.002, essentially
zero. Hours above 95 F average slightly *lower* demand (748 MW) than cooler hours
(751 MW), and the share of hours flagged as peak events sits near 10% in every
temperature band. Humidity and wind speed behave the same way. In this dataset,
weather does not move system demand, so a forecasting strategy built around
better weather feeds would be spending money in the wrong place.

# What Actually Moves Demand: Customer Segment Loads

```{r loads}
# Compare actual demand with the sum of the three customer segment loads
demand <- demand %>%
  mutate(LoadSum = ResidentialLoadMW + CommercialLoadMW + IndustrialLoadMW)

cor(demand$ActualDemandMW, demand$LoadSum)

# How the total load splits across segments
demand %>%
  summarise(Residential_Pct = round(sum(ResidentialLoadMW) / sum(LoadSum) * 100, 1),
            Commercial_Pct  = round(sum(CommercialLoadMW)  / sum(LoadSum) * 100, 1),
            Industrial_Pct  = round(sum(IndustrialLoadMW)  / sum(LoadSum) * 100, 1)) %>%
  kable(caption = "Share of total load by customer segment")

set.seed(42)
demand %>%
  slice_sample(n = 3000) %>%
  ggplot(aes(LoadSum, ActualDemandMW)) +
  geom_point(color = c_blue, alpha = 0.3, size = 1) +
  labs(title = "Sum of segment loads vs actual system demand",
       x = "Residential + Commercial + Industrial load (MW)",
       y = "Actual demand (MW)") +
  theme_nges()
```

Here is where the signal lives. The sum of residential, commercial and industrial
load correlates with actual demand at 0.984, and the residual between the two has
a standard deviation of only about 25 MW. Residential customers carry 46.6% of
the load, commercial 29.4% and industrial 24.0%. Demand spikes happen when two or
three segments run high at the same time, not when the weather turns. That points
Day 2 modeling squarely at segment-level load data from the AMI meters rather
than at weather forecasting.

# Peak Demand Events

```{r peaks}
# How do flagged peak events relate to the demand level itself?
demand %>%
  group_by(PeakDemandEvent) %>%
  summarise(Hours = n(),
            AvgDemand_MW = round(mean(ActualDemandMW), 1),
            MinDemand_MW = round(min(ActualDemandMW), 1),
            .groups = "drop") %>%
  kable(caption = "Peak vs non-peak hours")

# Peak-event rate by demand decile
demand %>%
  mutate(Decile = ntile(ActualDemandMW, 10)) %>%
  group_by(Decile) %>%
  summarise(PeakRate_Pct = round(mean(PeakDemandEvent) * 100, 1), .groups = "drop") %>%
  kable(caption = "Share of hours flagged as peak events, by demand decile")
```

Exactly 2,500 of the 25,000 hours (10%) are flagged as peak events, and the
decile table shows the flag is simply the top tenth of hourly demand. Peak hours
average 996.9 MW against 723.0 MW for everything else. This is useful because it
means "predicting peak events" and "predicting demand accurately at the high end"
are the same problem.

# Forecast Accuracy Today

```{r forecast}
# Overall accuracy of the existing NGES forecast
demand %>%
  filter(!is.na(ForecastedDemandMW)) %>%
  summarise(MAE_MW   = round(mean(abs(ActualDemandMW - ForecastedDemandMW)), 1),
            MAPE_Pct = round(mean(abs(ActualDemandMW - ForecastedDemandMW) /
                                  ActualDemandMW) * 100, 2),
            Bias_MW  = round(mean(ActualDemandMW - ForecastedDemandMW), 2)) %>%
  kable(caption = "Current forecast accuracy")

# Error distribution
demand %>%
  filter(!is.na(ForecastedDemandMW)) %>%
  ggplot(aes(ForecastErrorMW)) +
  geom_histogram(bins = 50, fill = c_blue) +
  labs(title = "Distribution of forecast error (actual minus forecast)",
       x = "Forecast error (MW)", y = "Hours") +
  theme_nges()

# Is the forecast worse during peak events?
demand %>%
  filter(!is.na(ForecastedDemandMW)) %>%
  group_by(PeakDemandEvent) %>%
  summarise(MAE_MW  = round(mean(abs(ActualDemandMW - ForecastedDemandMW)), 1),
            Bias_MW = round(mean(ActualDemandMW - ForecastedDemandMW), 2),
            .groups = "drop") %>%
  kable(caption = "Forecast error, peak vs non-peak hours")
```

The current forecast lands within 23.8 MW of actuals on average, a MAPE of about
3.3%, with essentially no overall bias (-0.04 MW). Interestingly, accuracy during
peak events (MAE 23.9 MW) is nearly identical to normal hours (23.8 MW), so the
client's Theory 3, that the forecast systematically underestimates peaks, is not
supported. The real issue is different: a plus-or-minus 24 MW miss matters far
more when the system is already at 997 MW and infrastructure is running near its
limits than it does on an average afternoon. Tightening that error band during
high-load hours is where forecasting value comes from.

# Infrastructure Risk & Utilization

```{r infra}
# Correlation of each factor with the composite risk score
infra %>%
  summarise(Age_vs_Risk      = round(cor(InfrastructureAgeYears, RiskScore), 2),
            Util_vs_Risk     = round(cor(CurrentUtilizationPct, RiskScore), 2),
            Outages_vs_Risk  = round(cor(OutagesLast5Years, RiskScore), 2),
            Growth_vs_Risk   = round(cor(ProjectedLoadGrowthPct, RiskScore), 2)) %>%
  kable(caption = "Correlation with substation risk score")

# Utilization bands
infra %>%
  mutate(UtilBand = cut(CurrentUtilizationPct, c(0, 70, 85, 90, 100, 999),
                        labels = c("Under 70%", "70-85%", "85-90%", "90-100%", "Over 100%"))) %>%
  count(UtilBand, name = "Substations") %>%
  ggplot(aes(UtilBand, Substations)) +
  geom_col(fill = c_blue, width = 0.55) +
  geom_text(aes(label = Substations), vjust = -0.4, fontface = "bold") +
  labs(title = "Substations by current utilization band",
       x = NULL, y = "Substations") +
  theme_nges()

# The ten highest-risk substations
infra %>%
  arrange(desc(RiskScore)) %>%
  head(10) %>%
  dplyr::select(SubstationID, Region, InfrastructureAgeYears, CurrentUtilizationPct,
         OutagesLast5Years, ProjectedLoadGrowthPct, RiskScore) %>%
  kable(caption = "Ten highest-risk substations")
```

The 500 substations average 26.9 years of age and 74.7% utilization, but the
distribution has a worrying tail: 127 substations run above 90% of rated
capacity and 35 are already above 100%. Age is the strongest single driver of the
risk score (r = 0.52), followed by utilization (0.44) and outage history (0.39).
Thirty-one substations score 70 or higher on risk, a group we treat as the
critical tier; the North (10) and West (8) regions hold the most of them. The
worst asset, substation 1103 in the South, combines 47 years of age with 8
outages in five years and 9.9% projected load growth. This table becomes the
starting point for Saturday's capital prioritization work.

# Demand Response Programs: First Look

```{r dr}
# Cost per MW of reduction is the natural effectiveness measure
dr <- dr %>% mutate(CostPerMW = ProgramCost / DemandReductionMW)

dr %>%
  group_by(CustomerType) %>%
  summarise(Events            = n(),
            AvgReduction_MW   = round(mean(DemandReductionMW), 1),
            AvgReduction_Pct  = round(mean(DemandReductionMW / PeakDemandBeforeMW) * 100, 1),
            MedianCostPerMW   = round(median(CostPerMW), 0),
            .groups = "drop") %>%
  kable(caption = "Demand response performance by customer type")

# The distribution of cost-effectiveness is the real story
ggplot(dr, aes(CostPerMW)) +
  geom_histogram(bins = 60, fill = c_blue) +
  scale_x_log10(labels = dollar) +
  labs(title = "Cost per MW reduced across 8,000 DR events (log scale)",
       x = "Cost per MW (log scale)", y = "Events") +
  theme_nges()
```

Across 8,000 events the program removed 1.33 million MW of peak load for $420.2
million. Customer types perform almost identically: residential, commercial and
industrial all cut about 11% of peak demand per event at a median cost near
$330/MW. The spread across individual events is enormous, though, from about $12
to $7,680 per MW. The program's problem is not which customers it recruits but
how inconsistently events are run. That cost-effectiveness tail is Saturday's
biggest financial opportunity, and we quantify it there.

# Key Findings & Day 2 Hypotheses

Five things stood out from Day 1:

1. Weather, calendar and region explain almost none of the variation in demand.
   Temperature correlates with demand at 0.002 and peak events occur at the same
   10% rate in every temperature band. Theory 1 is not supported by this data.
2. Demand is driven by customer segment loads (correlation 0.984 with the segment
   sum). Volatility comes from segments peaking together.
3. The current forecast (MAE 23.8 MW, MAPE 3.3%) is unbiased and no worse during
   peaks, but a 24 MW error band is expensive when 127 substations run above 90%
   utilization.
4. Infrastructure risk is concentrated: 31 critical substations, with age the
   strongest driver, and 35 assets already past 100% utilization.
5. Demand response works but is wildly inconsistent per dollar: identical median
   performance across customer types, with a long tail of events costing over
   $1,000/MW.

For Day 2 we will test three hypotheses: (a) a model built on segment loads will
beat the current forecast while a weather-based model will not; (b) a small
critical tier of substations should absorb most near-term capital; and (c)
rebalancing the least efficient quartile of DR events can recover most of the
$95-120M exposure leadership described. Data limitations to carry forward: 2%
missing weather and forecast values, impossible humidity readings, and the fact
that the demand file's 100 substations do not overlap cleanly with the
infrastructure file's 500, which keeps us from joining the two at asset level.
